---
title: "Time Series HW 1"
author: "Akshay Kumar Varanasi (av32826)"
date: "February 5, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Question 1**

a) Fit	an	AR(1)	model	to	the	data	using	methods	we	learnt	in	class.	Determine	
the	maximum	likelihood	estimate	of	the	autoregressive	coefficient$\phi_1$	?,	
variance	of	the	white	noise	in	the	AR(1)	model	and	the	variance	of	the	
maximum	likelihood	estimate of	the	parameter $\phi_1$.

```{r}

## To read the data file
library("xlsx")
data<-read.xlsx("HW1_Problem1.xls",sheetIndex = 1, header=FALSE, stringsAsFactors=FALSE)

## Hardcoding to find those parameters
sum=0
sum_sq=0
mu=mean(data$X1)
var_data=0

## Before going forward to do the model fitting, we need to subtract the data by its mean
data$X1<-data$X1-mu

## Calculating Phi_1
for (i in 1:49) {
     sum = sum + ((data$X1[i])*(data$X1[i+1]))
     sum_sq = sum_sq+((data$X1[i]))^2 #*(data$X1[i]))
}
phi_1=sum/sum_sq
print(paste("The value of phi_1 is",prettyNum(phi_1)))

## Calculating variance of white noise and phi_1
for (i in 1:49) {
  var_data=var_data+(data$X1[i+1]-phi_1*data$X1[i])^2
}
var_data=var_data/49
var_phi=var_data/sum_sq

print(paste("The value of variance of white noise is",prettyNum(var_data),"and that of phi_1 is",prettyNum(var_phi)))

## In built function to fit data in AR model with order 1
model<-ar.mle(x = data$X1, aic = FALSE, order.max = 1)

## Prints out co-efficient and variance value of white noise
model
```
```{r}

data$time<-1:length(data$X1)
data$predict<-rep(0,length(data$X1))
#data_predicted<-data_frame()
#data$predict<-predict(model, data$X1)

#data$predict<-predict(model, data$X1, n.ahead = 1, se.fit = TRUE)
for (i in 1:49) {
  data$predict[i+1]=data$X1[i]*phi_1
}

plot.ts(data$X1, lwd=2, col="Blue", xlab="Time", ylab="Data")
lines(x=data$time, y=data$predict, col='Red', lty=2 )  
legend("topright", legend=c("Original", "Predicted"),
       col=c("blue", "red"), lty=1:2, cex=0.8)
```

b) Comment	on	your	result?	What	kind	of	a	process	does	this	look	like?


The process which we predicted is more or less stationary around a constant due to low value of phi. This is because the values are not dependent on previous values, i.e the fluctuations or the trend in actual data is mostly due to noise. 

**Question 2**
For	the	retail	sale	data	posted	on	the	class	website	(originating	form	
before	the	economic	calamity	of	2007/2008),	please	do	the	following
(a) Fit	a	line	passing	through	the	data	(first	order	polynomial	fit).	Use	regression	
methods	used	in	class,	with	time	t being	indexed	from	0	to	length(data)-1	
(month	is	the	unit	of	time)

```{r}

## Reading the data
library("xlsx")
data<-read.xlsx("Retail_Sales_Data.xlsx",sheetIndex = 1, header=FALSE, stringsAsFactors=FALSE)

## Changing the column names for our convinience
colnames(data)<-c("Year","Month","Sales")

## Removing first 2 lines
data_sub<-data[3:length(data$Year), ]

## Converting the date and month to numbers
data_sub$Time<-c(1:length(data_sub$Sales))

## Creating a linear model relating Sales and time
model<-lm(data_sub$Sales~data_sub$Time)
summary(model)

## Plotting the model 
plot(fitted(model),residuals(model))

## Predicted values
pre_variables<-predict(model,data=data_sub$Time)

## Plotting both original and predicted values
plot.ts(data_sub$Sales, lwd=2, col="Blue", xlab="Time", ylab="Sales")
lines(x=data_sub$Time, lwd=2, y=pre_variables, col='Red', lty=2)
legend("topleft", legend=c("Original", "Predicted"),
       col=c("blue", "red"), lty=1:2, cex=0.8)



```


(b) Calculate	residuals	after	you	removed	the	first	order	fit	and	then	fit an	
autoregressive	model	of	order	2	– AR(2)	to	those residuals	(again,	months	
are	units	of	time). Please	mark	the	residual	sum	of	squares	after	you	fitted	
the	AR(2)	model.
```{r}

data_sub$Predicted<-pre_variables
data_sub$Resdiuals<-as.numeric(data_sub$Sales)-as.numeric(data_sub$Predicted)

# AR(2)
res<-ar.ols(x = data_sub$Resdiuals, aic = FALSE, order.max = 2)
res

# Residual Sum of squares
print(paste("The value of residual sum of squares for AR(2) is",prettyNum((length(data_sub$Year)-1)*res$var.pred)))

#data_sub$Resdiuals_pre<-predict(res,data_sub$Resdiuals)
#plot.ts(data_sub$Resdiuals)
#lines(x=data_sub$Time,y=data_sub$Resdiuals_pre)
```

(c) Take	the	same	residuals	obtained	after	fitting	a	line	through	the	sales	data,	
but	now	please	fit	an	AR(4)	model.	Once	again	mark	the	residual	sum	of	
squares	after	you	fitted	the	AR(4)	model.

```{r}

# AR(4)

res_4<-ar.ols(x = data_sub$Resdiuals, aic = FALSE, order.max = 4)
res_4
# Residual Sum of squares
print(paste("The value of residual sum of squares for AR(4) is",prettyNum((length(data_sub$Year)-1)*res_4$var.pred)))
```


(d) Increase	the	order	of	the	AR(k)	model	by	fitting	AR(6),	AR(8),	AR(10)	and	so	
on	models.	Do	this	until	you	fit	AR(20).	Please	plot	the	residual	sums	of	
squares	of	these	models	as	you	increased	the	order	of	the	AR	models.	Please	
comment	on	what	you	see

```{r}
res_sum_sq=rep(0L,10)
for (i in seq(2,20,2)){
  res<-ar.ols(x = data_sub$Resdiuals, aic = FALSE, order.max = i)

  res_sum_sq[i/2]=res$var.pred*(length(data_sub$Year)-1)
}
res_sum<-data.frame(res_sum_sq)
#res_sum$res_sum_sq<-res_sum_sq
res_sum$model_num=seq(2,20,2)

library(ggplot2)
ggplot(res_sum,aes(x=res_sum$model_num, y=res_sum$res_sum_sq))+geom_point(aes(color = 'Red'),show.legend = FALSE)+geom_line(aes(color = 'green'),show.legend = FALSE)+xlab("Order")+ylab("Residual sum of squares")
```



From the plot, we can see how residual sum of squares drops off suddenly after some increase in order number then stays around that value as we increase further. It is because of as we increase number of variables to fit the fit becomes better. 